---
title: "FIFA World Cup Analysis"
author: "Insert your name here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, 
                      include = TRUE,
                      fig.align = "center",  out.width = "80%")
```


## Introduction
In this study, our primary goal is to answer the question: "Which was the best World Cup?" To accomplish this, we make use of data derived from the FIFA World Cup, encompassing two primary datasets: **wcmatches** and **worldcups**. The **wcmatches** dataset compiles a comprehensive record of every World Cup match played between 1930 and 2018, featuring variables that encompass the `year`, `country`, `city`, participating teams, scores, match outcomes, and when the game took place. Notably, the variable `stage` provides insights into the tournament phase, including stages like the group stage, round of 16, quarter-finals, semi-finals, and the final. The `outcome` variable states the match results, with 'H' representing a home team win, 'A' for an away team win, and 'D' for a draw. Additionally, the `win_conditions` variable indicates special win conditions, such as extra time or penalties when not empty. The complementary **worldcups** dataset offers a summary of each World Cup, featuring crucial statistics, such as the `year`, `host` country, winning team, `goals_scored`, and `attendance` figures. This analysis aims to illuminate the factors contributing to the distinction of the "best" World Cup, taking into account various aspects. These aspects include the total number of goals scored, the frequency of games with special win conditions, and the geographical time zones where the matches were played. By exploring these datasets and their associated variables, we seek to provide insights into what defines an exceptional World Cup tournament.

### Data Download
```{r}
match_info = readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/wcmatches.csv',show_col_types = FALSE)
wc_info =  readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/worldcups.csv', show_col_types = FALSE)
```

## Data Cleaning and Summary
### Data Cleaning

It's worth noting that there are NA values present in the `win_conditions`, `winning_team`, and `losing_team` columns of the match_info table, and this is entirely normal. In instances where games concluded within the standard 90 minutes, the `win_conditions` column understandably remains NA, as no specific win conditions beyond regular play are applicable. Additionally, when matches result in a draw, both the `winning_team` and `losing_team` columns will be NA, as there is no distinct winning or losing team. This occurrence is reflected in the dataset, and it aligns with the nature of drawn matches where the scores for both teams remain equal.

#### match_info table:

```{r}
# Assuming your dataset is named my_data
na_counts_match <- colSums(is.na(match_info))
print(na_counts_match)

```
#### wc_info table:
```{r}
na_counts_wc <- colSums(is.na(wc_info))
print(na_counts_wc)
```
### Data Summary
```{r}
# Load the necessary libraries
library(dplyr)



# Number of World Cups
num_worldcups <- nrow(wc_info)

# Total goals scored
total_goals <- sum(wc_info$goals_scored)

# Extract unique cities from the host column
unique_cities_played <- unique(match_info$city)

# Number of unique cities where matches have been played
num_cities_played <- length(unique_cities_played)

# Extract unique countries from home_team and away_team columns
unique_countries <- unique(c(match_info$home_team, match_info$away_team))

# Number of unique countries that participated
num_countries_participated <- length(unique_countries)

# Display the summary statistics
cat("Summary Statistics for wc_info:\n")
cat("Number of World Cups: ", num_worldcups, "\n")
cat("Number of unique cities where matches have been played: ", num_cities_played, "\n")
cat("Total goals scored: ", total_goals, "\n")
cat("Number of unique countries that participated: ", num_countries_participated, "\n")

```

## Visualizations


### Win Rate vs. Goal Difference Scatter Plot

A compelling host country performance significantly contributes to the overall success and memorability of a World Cup. The host nation serves as the focal point of the tournament, and when they deliver an impressive performance on the field, it enhances the tournament's excitement and entertainment value. These exceptional performances generate a sense of national pride, fostering a vibrant and celebratory atmosphere throughout the host nation. Moreover, a host country's journey in the tournament can create enduring memories and iconic moments that define the World Cup experience. A successful host country performance, marked by skillful play, strategic victories, and passionate fan engagement, contributes to the narrative of a memorable and exceptional World Cup, making it a benchmark for future tournaments and ensuring its lasting impact in the history of football.

<br>

#### Variables:

- **Goal Difference (X-axis):** Represents the performance delta between goals scored and conceded (`home_score` - `away_score`). A positive goal difference indicates a strong offensive performance.

- **Win Rate (Y-axis):** Reflects the proportion of victories to total matches played. A higher win rate showcases the host country's success in match outcomes.

<br>

#### Interpretation:
- **Positive Goal Difference:** A positive value on the x-axis indicates that the host country scored more goals than they conceded, reflecting a strong offensive performance.

- **Higher Win Rate:** A higher point on the y-axis signifies a higher win rate, suggesting a successful overall performance in terms of match outcomes.

<br> 

#### Why scatter:

To visually explore the relationship between a host country's win rate and goal difference, a scatter plot is employed. A scatter plot is ideal for this analysis as it allows for the simultaneous observation of two continuous variables (win rate and goal difference) for each World Cup, facilitating the identification of patterns, clusters, and outliers in host country performances across different tournaments.

<br>

#### Note:
It's crucial to highlight that the 2002 World Cup was jointly hosted by Japan and Korea; therefore, each of these countries will be treated as a separate host nation.


```{r}
library(dplyr)
library(ggplot2)

# Filter matches where country = home_team and add a new column year1
host_match_info <- match_info %>%
  filter(country == home_team) %>%
  mutate(
    year1 = case_when(
      year == 2002 & country == "South Korea" ~ "2002_KR",
      year == 2002 & country == "Japan" ~ "2002_JP",
      TRUE ~ as.character(year)
    )
  )

# Count the number of 'H' and the number of matches for each year1
host_win_rate <- host_match_info %>%
  group_by(year1) %>%
  summarize(
    total_matches = n(),
    total_wins = sum(outcome == 'H'),
    host_win_rate = total_wins / total_matches
  )

# Create a table with goal difference (sum of home_score - sum of away_score) for each year1
goal_difference <- host_match_info %>%
  group_by(year1) %>%
  summarize(goal_difference = sum(home_score) - sum(away_score))

# Join the goal_difference table and host_win_rate table
scatter_data <- left_join(host_win_rate, goal_difference, by = "year1")

# Plot a scatter of win_rate and goal_difference with labels
ggplot(scatter_data, aes(x = goal_difference, y = host_win_rate, color = year1)) +
  geom_point() +
  geom_text(aes(label = year1), hjust = 0.5, vjust = -0.5, size = 3) +
  labs(
    title = "Scatter Plot of Win Rate vs. Goal Difference for Host Country",
    x = "Goal Difference",
    y = "Win Rate",
    color = "Year"
  ) +
  theme_minimal() +
  guides(color = FALSE)  # Remove the legend

```

### Diverging Bar Chart on Goals Scored Per Game in Each World Cup
More goals often contribute to the excitement and entertainment of the tournament and a high-scoring tournament often create lasting memories.Goals are the emotional currency of football. The elation of witnessing your team score and the heartbreak of conceding create a roller coaster of emotions for fans. A World Cup with more goals offers fans numerous opportunities to experience the collective highs and lows that make football a deeply emotional and immersive experience.Fans remember the drama, suspense, and emotion associated with the goals, and these moments become an integral part of the World Cup's historical legacy.

<br>

#### Variables:

- **Year (X-axis):** Represents the year of the World Cup

- **Difference Between Average and Goals Scored Per Game (Y-axis):** Represents the magnitude of difference from the average goals scored per game.

<br>

#### Why bar chart:
 This type of visualization is ideal for comparing the goal-scoring performance of different World Cups as it provides a clear overview and allows for easy identification of tournaments with the highest average goal per game counts.

```{r}
# Load the required libraries
library(dplyr)
library(ggplot2)

# Load the data
match_info = readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/wcmatches.csv',show_col_types = FALSE)
wc_info =  readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-29/worldcups.csv', show_col_types = FALSE)

# Summarize goals per game for each World Cup
total_goals_per_worldcup <- match_info %>%
  left_join(wc_info, by = "year") %>%
  group_by(year) %>%
  mutate(goals_per_game = sum(home_score + away_score) / games, year = as.factor(year)) %>%
  select(year,goals_per_game) %>%
  unique() %>%
  ungroup() %>%
  mutate(divergence = goals_per_game - mean(goals_per_game))

# Create a bar chart
ggplot(total_goals_per_worldcup, aes(x = reorder(year,divergence),y = divergence)) +
  geom_bar(stat = "identity", color = "white", size = 0.7, width = 0.7, position = position_dodge(width = 0.8), aes(fill = divergence > 0)) + 
  scale_fill_manual(labels = c("Below Average","Above Average"),
                    values = c("FALSE" = "#f8766d", "TRUE" = "#00ba38"),
                    name = "Goals Scored Per Game") +
  coord_flip() + 
  labs(title = "Goals Scored Per Game in Each World Cup",
       subtitle = "vs Average Goals Scored Per Game ",
       x = "",
       y = "",
       fill = "Goals Scored Per Game > Average") +
  theme_bw() +
  theme(axis.title = element_text(size = 12, face = "bold", color = "black"),
        plot.title = element_text(size = 14, face = "bold", color = "black"),
        legend.key.size = unit(0.5, "cm"),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8),
        legend.position = "bottom",
        panel.border = element_blank(),  # Remove panel border
        panel.background = element_blank(),  # Remove panel background
        plot.background = element_rect(fill = "#f5f5f5", color = NA),  # Set plot background color
        plot.margin = unit(c(1, 1, 1, 1), "cm"))  # Adjust plot margins

```

### Stacked Area Chart: Games with Special Win Conditions
Understanding the dynamics of games with special win conditions, such as extra time or penalties, is pivotal in unraveling the excitement and intensity of a World Cup tournament. These moments, exclusive to knockout stages, are characterized by heightened stakes, adding a layer of unpredictability and thrill to the matches. As these special conditions unfold in later stages of the competition, the tournament's overall spectacle is elevated, contributing to the tournament's perceived greatness and memorability. This analysis seeks to shed light on the frequency of these high-stakes encounters across different World Cup years.

<br>

#### Variables:
- **Year of World Cup (X-axis)**: Represents the chronological order of World Cup tournaments from 1930 to 2018.

- **Percentage of Games with Special Win Conditions (Y-axis)**: Reflects the proportion of matches where special win conditions, such as extra time or penalties, were applied, relative to the total number of matches in a given World Cup year.

<br>

#### Additional Details:
- Justification for Using Percentage: The decision to represent the data in terms of percentages instead of counts for each special win type is grounded in the understanding that different tournaments have varying numbers of games. By using percentages, the analysis accounts for these variations, providing a more nuanced perspective on the prevalence of special win conditions in each World Cup. This approach ensures a fair comparison across tournaments with differing match counts and better captures the impact of these high-stakes moments on the overall tournament experience.

<br>

#### Why Stacked Area Chart:
A stacked area chart is an effective choice for visualizing the prevalence of special win conditions across different World Cup years. This chart allows for a clear comparison between the two types of special win conditions, showcasing their individual contributions to the overall tournament narrative. The stacked format emphasizes the cumulative impact of these moments, providing a comprehensive view of their occurrence and distribution.

<br>

#### Notes:
- There is no World Cup tournament in 1942 and 1946 due to World War II.
- Penalties were first introduced in the 1978 World Cup tournament.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)

# Assuming wc_info contains the columns year and games
# Identify games with special win conditions
special_win_conditions <- match_info %>%
  filter(!is.na(win_conditions)) %>%
  mutate(win_type = ifelse(grepl("AET", win_conditions), "Extra Time", "Penalties"))

# Create a new dataframe with all years
all_years <- data.frame(year = seq(1930, 2018, by = 4))

# Count the number of games with special win conditions for each type
special_win_counts <- special_win_conditions %>%
  group_by(year, win_type) %>%
  summarize(count = n()) %>%
  spread(win_type, count, fill = 0) %>%
  right_join(all_years, by = "year") %>%
  replace(is.na(.), 0)

# Merge with wc_info to get the total number of games for each year
special_win_counts <- special_win_counts %>%
  left_join(wc_info, by = "year")

# Calculate the percentage of games with special win conditions
special_win_counts <- special_win_counts %>%
  mutate(
    extra_time_percentage = `Extra Time` / games * 100,
    penalties_percentage = Penalties / games * 100
  )

# Reshape data for stacked area chart
stacked_area_data <- special_win_counts %>%
  gather(key = "win_type", value = "percentage", extra_time_percentage, penalties_percentage) %>%
  select(-`Extra Time`, -Penalties)

# Replace NA with 0 in the percentage column of stacked area data
stacked_area_data$percentage[is.na(stacked_area_data$percentage)] <- 0

# Create a stacked area chart with viridis color scale and theme_ipsum
ggplot(stacked_area_data, aes(x = as.factor(year), y = percentage, fill = win_type, group = win_type)) +
  geom_area(position = "stack", show.legend = TRUE, alpha = 0.7, size = 0.5, color = "white") +
  labs(
    x = "Year of World Cup",
    y = "Percentage of Games with Special Win Conditions",
    title = "Stacked Area Chart of Games with Extra Time and Penalties in Each World Cup",
    fill = "Win Type"
  ) +
  scale_fill_viridis(discrete = TRUE, option = "magma") +
  theme(legend.position = "top")
```

